---
title: "Session 5: Crash course in data wrangling"
author: "Heiner Salomon"
institute: "Bremen University"
date: "18 05 2026"
format:
  beamer:
    theme: "Berlin"
    colortheme: "beaver"
    incremental: true
    header-includes:
    - \apptocmd{\tightlist}{\setlength{\itemsep}{20pt}}{}{}
    - \logo{\includegraphics[height=1cm]{UniBremen_Logo_Pantone186.pdf}}
toc: true
toc-title: "Overview"

echo: TRUE
eval: FALSE
---

# Introduction

## Learning goals of today

-   Short repetition from last week
-   Discussing "replication crisis" and potential solutions
-   Crash course in data wrangling

## Repetition: Useful commands

-   mutate() \[+ case_when\]: Change data and create new variables
-   filter(): Filtering data by specific values
-   select(): Selecting variables

## Repetition: Piping in tidyverse

```{r}
# "Piped" code (as in the tidyverse)
iris_data <- iris |>
  filter(Species = "virginica") |>
  mutate(Petal.Length = Petal.Length + 1)
```

## Reading - Reproducible science

-   What is p-hacking
-   What is a pre-analysis plan?
-   What is publication bias?
-   How can bias creep into your research?

## Reading - Suggestions they provide

-   Improving methodological training
-   Encouraging collaboration and team science
-   Promoting study pre-registration
-   Improving the quality of reporting
-   Promoting transparency and open science (data sharing)

# Data wrangling

## What will we learn today

-   Some more data wrangling commands
-   group_by()
-   join()
-   arrange()
-   pivot()

## Why do we need to learn data wrangling?

-   Most of your time you will spend data wrangling
-   Analysis itself will take less than 10% of your time
-   Goal: To create clean, "tidy" data

## What is tidy data?

- Each variable is a column; each column is a variable.
- Each observation is a row; each row is an observation.
- Each value is a cell; each cell is a single value.
-   For more information on tidy data: https://tidyr.tidyverse.org/articles/tidy-data.html



## group_by() - Applying manipulation by groups of data

-   Often you will need to manipulate data grouped by another variable
-   Example: You have individuals, but you want to know how many people live in each city
-   group_by() can be inserted before mutate operation in pipe
-   group_by() manipulates data until ungroup() called

## group_by() - Variable manipulation (code)

```{r}
# Create sums by Species
iris_data <- iris |> 
  group_by(Species) |>
  mutate(sum_Sepal.Length = sum(Sepal.Length)) |>
  ungroup() # always (!) remove grouping
```

## .by = Alternative to group_by()

-   Instead of having to call group_by() and worry about ungroup()
-   Alternative: add .by = at the end of tidyverse command
-   Only operates on this one function
-   I prefer .by = over group_by()

## .by = Alternative to group_by() (code)

```{r}
# Create sums by Species
iris_data <- iris |> 
  mutate(sum_Sepal.Length = sum(Sepal.Length), .by = Species)
```

## Test: group_by() and .by = manipulate a variable by two groups and

-   load your data sets
-   use mutate() together with group_by() or .by 0 to create two new variables

## join() - Combining different data sets

-   join() commands necessary for comibining data sets
-   requires key variable to join on in both data sets!
-   left_join(a,b) join b onto a, most common version
-   others are right_join(), inner_join(), semi_join(), anti_join(), full_join()

## join() - Combining different data sets - join_by()

-   key variable(s) need to be uniquely identify
-   add join_by() after comma behind left_join()
-   join_by(x == y) will join data set a and b using variable x from data set a and y from data set b
-   Multiple join_by(x == y, h == j) possible

## join() - Combining different data sets - type of join

-   different kinds of joins possible
-   one-to-one, many-to-one, one-to-many, many-to-many (the latter hopefully never!)
-   relationship = "one-to-one" after comma behind left_join() to prevent errors

## join() - Combining different data sets (code)

```{r}
# Check whether keys identify uniquely (duplicated(sum) has to be 0!)
sum(duplicated(band_members_data$name))
sum(duplicated(band_members$name))

# Create sums by Species in band_members_data
band_members_data <- band_members |>
  left_join(band_instruments, join_by("name" == "name"), relationship = "one-to-one")
```

## Test: join() - Combining different data sets (code)

-   Do you have multiple data sets you would want to use?
-   use left_join() together with join_by() to create two new variables

## arrange() - sort (multiple) variables

-   arrange() sorts variables from highest to lowest
-   can be inversed from lowest (top) to highest using desc()
-   can take multiple variables

## arrange() - sort (multiple) variables (code)

```{r}
# Sort entries by Petal.Width and Sepal.Width
iris_data <- iris |>
  arrange(Petal.Width, Sepal.Width)

View(iris_data)

```

## Test: arrange() -  sort (multiple) variables

-   Try to sort your data set using first one, then two, then three variables


## pivot() - "switch columns with rows"

-   Often data will not come in tidy form
-   there are two pivot options
-   pivot_longer() - moves data from columns into fewer variables with more rows
-   pivot_wider() - moves data from rows into more variables with fewer rows

## pivot_longer() - example - before...

| Name     | Week1 | Week2 | Week3 | Week4 |
|----------|-------|-------|-------|-------|
| Caroline | 2     | 3     | 6     | 3     |
| Jack     | 5     | 6     | 7     | 5     |
| Ariana   | 3     | 4     | 3     | 7     |

## pivot_longer() - example - ...and after

| Name     | Week | Amount |
|----------|------|--------|
| Caroline | 1    | 2      |
| Caroline | 2    | 3      |
| Caroline | 3    | 6      |
| Caroline | 4    | 3      |
| Jack     | 1    | 5      |
| Jack     | 2    | 6      |
| Jack     | 3    | 7      |
| Jack     | 4    | 5      |
| Ariana   | 1    | 3      |
| Ariana   | 2    | 4      |
| Ariana   | 3    | 3      |
| Ariana   | 4    | 7      |

## pivot_wider() - example - before...

| Name     | Week | Amount |
|----------|------|--------|
| Caroline | 1    | 2      |
| Caroline | 2    | 3      |
| Caroline | 3    | 6      |
| Caroline | 4    | 3      |
| Jack     | 1    | 5      |
| Jack     | 2    | 6      |
| Jack     | 3    | 7      |
| Jack     | 4    | 5      |
| Ariana   | 1    | 3      |
| Ariana   | 2    | 4      |
| Ariana   | 3    | 3      |
| Ariana   | 4    | 7      |

## pivot_wider() - example - ...and after

| Week | Caroline | Jack | Ariana |
|------|----------|------|--------|
| 1    | 2        | 5    | 3      |
| 2    | 3        | 6    | 4      |
| 3    | 6        | 7    | 3      |
| 4    | 3        | 5    | 7      |

## pivot() - "switch columns with rows"

- pivot can be quite complicated
- for more information, see ?pivot_longer() or ?pivot_wider

# Any questions?

# Working on projects
