---
title: "Session 3: Introduction to R in Rstudio"
author: "Heiner Salomon"
institute: "Bremen University"
date: "04 05 2026"
format:
  beamer:
    theme: "Berlin"
    colortheme: "beaver"
    incremental: true
    header-includes:
    - \apptocmd{\tightlist}{\setlength{\itemsep}{20pt}}{}{}
    - \logo{\includegraphics[height=1cm]{UniBremen_Logo_Pantone186.pdf}}
toc: true
toc-title: "Overview"

echo: TRUE
eval: FALSE
---

# Introduction

## Learning goals of today

-   Short repetition from last week
-   Talking about best coding practices
-   Discussing research questions
-   Introduction to R in Rstudio

# Short repetition from last week

## Project

-   Projects are useful "shells" for your work
-   Make your work portable
-   Simplifies pathing within projects

## Rstudio

-   Four different windows
    -   Script (top left)
    -   Temporary files ("Environment", top right)
    -   R - Console (using R interactively, bottom left)
    -   Files/ Plots/ Help (bottom right)

## Quarto

-   Introduction to header for Quarto projects (YAML)
-   Basics of file system

# Short discussion of tasks

## Reading

-   Did you find the text useful?

-   What is your main takeaway?

## Research question

-   Please state your research question

-   What data (and variables) do you want to use?

-   Current questions?

# RStudio basics

## What will we learn

-   Today only basics (!)
-   What is what in R
-   R basics
-   Packages and help
-   Loading data
-   Summarizing data

## What will we learn next session (tidyverse)

-   Mutating variables
-   Merging data

## Basic coding best practices

-   keep it simple
-   "The computer" is stupid, so spelling matters (!)
-   try to avoid spaces (in names!) and upper-case letters
-   in R indentation does not matter, in Quarto it can

## Terminology: What is what in R?

-   Object \<- can be "anything"
-   Functions are a (set of) operations that do something specific to an object
-   Data frame \<- tabular data (rows and columns) loaded into R
-   Value \<- something that has been assigned to a name
-   List \<- number of different objects combined

## Terminology: Data types in R?

-   Numeric \<- any variable with free numbers
    -   eg population, meters etc.
-   Factor \<- specific number with specific meanings
    -   eg sex, employment status etc.
-   String \<- any text, delineated by ""

## Assignment

-   Assigning an object to name to refer to it later
-   Using the \< followed by - like this "\<-"
-   Gets stored in top right under "Global environment"

```{r}
# Load a data set
iris_data <- iris
# Assign value 2 to x
x <- 2
```

## How to get to variables in Base R?

-   We will mostly use tidyverse
-   still this seemed quite relevant
-   refer to next level of nesting using \$
-   is visualised through clickable errors in "Environment"

```{r}
# Speichern der Variablen Sepal.Length von iris in y
y <- iris$Sepal.Length

```

## Commenting

-   Comments are essential in understanding code
-   Comments are added by starting line with \#

```{r}
# This is a comment
2+2
```

## Code sections

-   You should try to code in sections
-   Like headers in text, makes it easier to follow code
-   Sections let you "jump" to code in outline and at bottom of script window

## Code sections - how to

-   Consist of comment \# followed by section name followed by at least four - - - -
-   Hidden ("folded") using arrow on left next to line number
-   Un-hidden using the blue double-arrow next to section name

```{r}
# This is a section ----
2+2
```

## Code sections - subsections

-   Can be useful at times in longer files
-   You can use \## or \### for different subsections

```{r}
# This is a section ----
v <- 2
## This is a sub-section ----
w = v + 3

```

## Packages

-   R is built up by a whole ecosystem of "packages"
-   packages are essentially a (set of) function(s) others wrote to simplify your life
-   "Under the hood" is the so-called Base-R
-   But most people use packages, especially tidyverse

## Packages - how to

-   First package needs to be installed using "install.packages("")"
-   Then loaded using "library()"

```{r}
# Install the tidyverse
install.packages("tidyverse")
install.packages("tidylog")
# Load the package
library(tidyverse)
library(tidylog)
```

## Packages - life-hack

-   usually, one loads all the necessary packages at the beginning of a script
-   if you don't want to do that you can also use the following
-   packagename::functionname

```{r}
# Using a package without loading it
readr::read_csv("your_data.csv")

```

## Help!

-   if you need help with a function, type in ? followed by the function name
-   Many tidyverse functions have own website
-   alternative: throw your question in your search machine

```{r}
# Help me with the function summary()
?summary()
# Then you can use it
summary(iris)
```

## Loading data

-   loading data into R using load()
-   often data not stored in Rdata file
-   import from other file types necessary

## Importing data - useful packages

-   Excel (.xlsx) -\> readxl::read_excel()
-   Delimited (.csv) -\> readr::read_csv()
-   Stata (.dta) -\> haven::read_dta()
-   SPSS (.sav) -\> haven::read_sav()

## Summarizing data

-   It is crucial (!!!) to understand the data one is working with
-   Many functions can be used to get overview

## Summarizing data - View()

-   View(name_dataframe) to see data
-   Alternative to clicking on data set in "Environment" on top right
-   opens up new tab with spreadsheet view on data
-   can be very slow and unwieldy with large data sets

## Summarizing data - head()

-   head(name_dataframe) to see first five rows of data
-   good quick overview
-   much faster than View()
-   May not show data complexity

## Summarizing data - names()

-   names(name_dataframe) to see list of all variable names
-   can be quite useful
-   not always helpful in long data sets

## Summarizing data - summary()

-   summary(name_dataframe) to see distribution of variables
-   good quick overview
-   limited usefulness, I personally find

## Summarizing data - table()

-   table(name_dataframe\$variable_name) to see distribution of one variable
-   table(name_dataframe\$variable1_name, name_dataframe\$variable2_name) to see cross-tabulation of two variables
- only shows counts, no percentages

## Summarizing data - skimr::skim()

-   install skimr
-   library(skimr) and skim(name_dataframe) or skimr::skim() to see data set
-   split up by type of variable
-   very good quick overview
-   can be quite slow in large data sets

# Break

# Tasks for the week

## Tasks for the week

-   No specific tasks this week
-   keep on working on your projects

## Reading for next week

-   Baker, D. (2013): How Much Unemployment Was Caused by Reinhart and Rogoff's Arithmetic Mistake? <https://cepr.net/publications/how-much-unemployment-was-caused-by-reinhart-and-rogoffs-arithmetic-mistake/>
-   We will discuss those ideas at the beginning of next week's class

# Any questions?
