---
title: "Session 4: Introduction to the tidyverse"
author: "Heiner Salomon"
institute: "Bremen University"
date: "11 05 2026"
format:
  beamer:
    theme: "Berlin"
    colortheme: "beaver"
    incremental: true
    header-includes:
    - \apptocmd{\tightlist}{\setlength{\itemsep}{20pt}}{}{}
    - \logo{\includegraphics[height=1cm]{UniBremen_Logo_Pantone186.pdf}}
toc: true
toc-title: "Overview"

echo: TRUE
eval: FALSE
---

# Introduction

## Learning goals of today

-   Short repetition from last week
-   Discussing Reinhart-Rogoff
-   Introduction to tidyverse

## Repetition

- Assignment: \<-
-   Commenting: \#
-   Sections: \# ----

## Repetition: Useful commands

-   Importing data packages: haven (Stata, SPSS), readxl (Excel), readr (CSV)
-   Looking at your data: View()
-   Summarizing data: skimr::skim()

## Reading - Reinhart-Rogoff

-   How did the mistake come about?
-   What are potential ways to prevent you doing similar mistakes?

# Tidyverse

## What will we learn today

-   What is the tidyverse?
-   "Piping"
-   mutate()
-   filter()
-   select()


## What is the tidyverse?

- tidyverse is many packages useful for typical data analysis
- Built on base R, tries to be more legible
- Important packages (for me) are: ggplot2, dplyr, tidyr, stringr...
- For more information: https://tidyverse.org/

## Before we continue - tidylog()

- tidyverse commands very "quiet"
- too easy to miss important changes to data
- tidylog() changes that and is extremely useful
- Add tidylog at end of beginning of each script packages loaded in library()

## "Piping" - the most important concept of the tidyverse

-   One of the most important concepts
-   A pipe is an operator: \|> (or sometimes %>%)
-   Follows the logic of steps (first step \|> second step \|> third step)
- usually starts with data set followed by function

## Functions - basics

- Short detour on functions
- Functions in R comprised of different "arguments" separated by \,
- Order of arguments matters (or have to specify arguments)

```{r}
# Understand standard arguments for head()
?head()

# Identical operations
head(iris, 10) # Order determines arguments
head(n = 10, x = iris) # arguments specified
```

## "Piping" - relation to functions

- In Base R, nesting of functions visually very confusing
- Piping replaces first argument in next function with last step in piping process
- As result, "piped" code easier to understand and fix


## "Piping" - a visual explanation

- Piping is adding each former step as first argument in next function
```{r}
# "Piped" code (as in the tidyverse)
iris_data <- iris |>
  filter(Species = "virginica") |>
  mutate(Petal.Length = Petal.Length + 1)

# Spelled out in brackets (as in Base R)
iris_data <- mutate( filter(iris, Species = "virginica"),
                     Petal.Length = Petal.Length + 1)

```


## "Piping" - an example

```{r}
# Step 1: Load a data set
iris_data <- iris
# Step 2: Filter the data
iris_data <- filter(iris_data, Species = "virginica")
# Step 3: Add one to each Petal.Length 
iris_data <- mutate(iris_data, Petal.Length =
                      Petal.Length + 1)

# All in one go
iris_data <- iris |>
  filter(Species = "virginica") |>
  mutate(Petal.Length = Petal.Length + 1)

```


## mutate() - Variable manipulation

-   Often you will need to create new variables or change existing ones
- Create new variable: mutate(data, new_variable = old_variable)
- old_variable and new_variable can be the same
- At the right between "old_variable" and the \")", you can add anything to manipulate the relationship


## mutate() - Variable manipulation (code)

```{r}
# Check the summary of Petal.Length before
iris_data |>
  select(Petal.Length) |>
  summary()

# Add 1 to each value of Petal.Length
iri_data <- iris |>
  mutate(Petal.Length = Petal.Length + 1)

# Check the summary of Petal.Length after
iris_data |>
  select(Petal.Length) |>
  summary()
```

## Test: mutate() - Variable manipulation

- load your data sets
- use mutate() to create a new variable

## mutate() - Conditional manipulation

- Often you only need conditional manipulation
- Can be achieved combining mutate() with case_when()
- case_when(condition1 ~ new_value1, condition2 ~ new_value2,...)

## mutate() - Conditional manipulation (code)

```{r}
# Check the distribution of Petal.Length before
iris |>
  select(Petal.Length) |>
  table()
# Using a simple binary case_when()
iris_data <- iris |>
  mutate(Petal.Length.binary =
           case_when(
             Petal.Length < 2 ~ "Small",
             Petal.Length >= 2 ~ "Large"
             ))
# Check the distribution of Petal.Length after
iris |>
  select(Petal.Length) |>
  table()
```

## Test: mutate() - Conditional manipulation

- use mutate() with case_when() to create a new variable based on conditions found in another variable


## filter() - Filtering "rows"

- Often data sets contain a lot of data
- too much data might slow down operations and be confusing
- filter() filters out unnecessary rows
- requires the \=\=, not only one \=
- Usually, the \! indicates a reversal

## filter() - Filtering "rows" (code)

```{r}
# Check the number of rows before
nrow(iris_data)

# Filter out all species not "Virginica"
iris_data <- iris |>
  filter(Species == "virginica")

# Check thenumber of rows after
nrow(iris_data)

```

## select() - Filtering "columns"

- select() is the same as filter(), but for columns (variables)
- uses "tidy selection", such as \:, c() etc.
- for more information, check: https://dplyr.tidyverse.org/reference/select.html
- Usually, the \! indicates a reversal


## select() - Filtering "columns" (code)

```{r}
# Check the variables before
names(iris_data)

# Select all variables in between Petal.Length and Species and keep them
iris_data <- iris |>
  select(Petal.Lentgh:Species)

# Check the variables after
names(iris_data)

```

## Test: select() - Filtering "columns"

- use select() to reduce the number of columns in your data set
- please make sure to not assign the old data set name to not "lose" the data



## Code formatting

-   Code formatting seems tedious
-   But formatting is important for clarity
-   Formatting speaks to line breaks, indentation and spacing
-   For tidyverse style guide, see https://style.tidyverse.org/

## Code formatting - using styler()

-   You can also use a package for formatting, styler()
-   It also includes a RStudio plugin
-   You can find it here: https://styler.r-lib.org/

## Some useful hotkey tips

-   RStudio has several very useful hotkeys, for instance:
-   Bulk - commenting: "Comment out" larger chunks of R-code -\> Ctrl + Shift + C
-   Multi-line cursor: Change code in multiple lines simoultaneously -\> Click beginning line + Hold Alt + Shift and click end line

# Tasks for the week

## Tasks for the week

- Please don't forget to send me progress updates by Friday 15:00
- This time we will use email

## Reading for next week

Munafò, Marcus R., Brian A. Nosek, Dorothy V. M. Bishop, et al. ‘A Manifesto for Reproducible Science’. Nature Human Behaviour 1, no. 1 (2017): 0021. https://doi.org/10.1038/s41562-016-0021
-   We will discuss those ideas at the beginning of next week's class

# Any questions?
